Management induced changes of soil organic carbon on global croplands

### 2022-07-12 SOCbudgetPaper/discussion.Rmd
update

For most of the parameters used in our management estimates no uncertainty estimates exist. This is why, in our view, most of the uncertainty with respect to the impacts of SOC management is included in the management data itself, and especially in the residue and manure production and application numbers, as these are only indirectly derived from crop and livestock production, feed and area data [@faostat_faostat_2016; @weindl_livestock_2017]. The uncertainty of recycling shares adds on top of the uncertain total numbers of manure and residue biomass. Previous modeling studies of SOC carbon on cropland often only used stylized scenarios of management practices [@pugh_simulated_2015; @lutz_simulating_2019], rather than trying to estimate real management.

@@

The Tier 2 approach [@ogle_cropland_in_ipcc_2019] used here is explicitly designed for agricultural soils, whereas we apply it also to soils under PNV. This is necessary in order to represent SOC losses under land-use change in a dynamic way, as this is an important driver of SOC dynamics. The comparison of simulated PNV data with LPJmL4 shows the model's substantial capability in reproducing PNV SOC stocks (Fig. \@ref(fig:SOCLPJmL4)).

Using litterfall estimates from LPJmL4, we have been able to estimate the total SOC stocks of the world, which is dominated by SOC under natural vegetation. However, as the world's SOC stock is highly uncertain, which is seen in the wide range of global SOC stock estimates for the first 30 cm of the soil profile [@batjes_harmonized_2016; @hengl_soilgrids250m_2017; @fao_global_2018; @schaphoff_lpjml4_2018-1; @poggio_soilgrids_2021; @sanderman_soil_2017] in Fig. \@ref(fig:SOCglo), the only conclusion we can draw from this is that our result is within a plausible range.
To avoid a strong impact of natural land representation and its uncertainties on our results, we focus on SOC changes on cropland. Pristine natural vegetated areas (like permafrost and rain forests) without human land management drop out in our calculation of SOC debt and stock change factors. Natural SOC estimates only influence results when natural land is converted to cropland.

@@

In our model, erosion should however only affect the spatial pattern but not the aggregate SOC pool. As pointed out by Doetterl et al. [-@doetterl_erosion_2016], the final fate of leached or eroded carbon is uncertain and might even offset LUC emissions [@wang_human-induced_2017]. Whereas for soil quality analysis SOC displacement might play an important role, in this budget approach focused especially on the SOC debt, displaced but not emitted SOC can be treated as SOC that remains on the cropland. Erosion and degradation impacts on yields and therefore on soil C inputs are captured by our method as we base them on FAO statistics of actual production. Yet the distribution of production below the country level - which we allocate proportional to LPJmL production potentials that do not reflect erosion feedback to yields - will overestimate yields and therefore biomass inputs to eroded soils.

In comparison with default stock change factors of the IPCC guidelines, our model estimates a stronger decline of SOC stocks (Table \@ref(tab:SCFglo)) for almost all regions. Tropical soils might suffer from low C input rates due to large yield gaps (Global Yield Gap and Water Productivity Atlas. Available URL: www.yieldgap.org (accessed on: 03/01/2022)) and high shares of residue removal and burning in lower-income countries [@smil_crop_1999; @williams_influence_1997; @jain_emission_2014]. Yet, even when comparing our etimates to the low-input stock change factors of the IPCC, our SOC loss is roughly twice as large as the revised 2019 IPCC default values, while it shows very good agreement with the older default values from IPCC [-@lasco_cropland_in_ipcc_2006]. Don et al. [-@don_impact_2011] estimated SOC losses for tropical soils of around 25% on average corresponding to a stock change factor of 0.75, but also reported a wide range of measured SOC changes from -80% to +58%. Fujisaki et al. [-@fujisaki_forest_2015] however found much lower loss rates of around 9%, attributing the difference to the different time period length since the conversion to cropland. As our results do not specifically account for cropland age and most of the cropland is older than 20 years (as assumed for the default IPCC Tier 1 stock change factors) our stock change factors have to be lower by definition following the steady-state assumption that cropland will continue to approach a new equilibrium. For the same reason, our estimates for temperate regions might be lower than both IPCC [-@lasco_cropland_in_ipcc_2006] and IPCC [-@ogle_cropland_in_ipcc_2019] default values.
With the production-increasing impact of irrigation and fertilization on carbon-poor dryland soils, SOC under cropland can also be higher than under PNV with stock change factors above 1 (see Fig. \@ref(fig:SOCmaps)(d)), but these areas are much smaller than where the stock change factors are well below unity.

@@

The effects of agricultural productivity on cropland SOC dynamics, including historical yield trends and associated increases in residue inputs, can be directly accounted for in our modeling approach. In contrast, process-based studies [@pugh_simulated_2015; @herzfeld_soc_2021] often lack data on relevant management aspects that drive production increases. Herzfeld et al. [-@herzfeld_soc_2021] also consider historical management trends for fertilizer and manure inputs as well as on residue removal rates and tillage systems, but cannot reproduce the substantial increase in agricultural productivity over the last decades. Still, they find that compared to no-tillage systems, residue management has much larger potential to affect the strength of their projected future global cropland SOC decline. This is consistent with our finding that increasing SOC inputs from above-ground residues had the strongest effect on the slowing-down of the SOC debt increase (Fig. \@ref(fig:SOCscen)).

>>>>

For most of the parameters used in our management estimates no uncertainty estimates exist. This is why, in our view, a large part of the uncertainty with respect to the impacts of SOC management --- next to the parametric and structural uncertainty of the soil model --- is included in the management data itself. This is especially the case for the residue and manure production and application numbers, as these are only indirectly derived from crop and livestock production, feed and area data [@faostat_faostat_2016; @weindl_livestock_2017]. The uncertainty of recycling shares adds on top of the uncertain total numbers of manure and residue biomass. Previous modeling studies of SOC carbon on cropland often only used stylized scenarios of management practices [@pugh_simulated_2015; @lutz_simulating_2019], rather than trying to estimate real management.

@@

The Tier 2 approach [@ogle_cropland_in_ipcc_2019] used here is explicitly designed for agricultural soils, whereas we apply it also to soils under PNV. This is necessary in order to represent SOC losses under land-use change in a dynamic way, as this is an important driver of SOC dynamics. The comparison of simulated PNV data with LPJmL4 shows the model's capability in reproducing PNV SOC stocks (Fig. \@ref(fig:SOCLPJmL4)).
Concurrently, the point-data comparison (see Fig. \@ref(fig:SOCpoint)) shows low correlation for PNV, however also for cropland sites. This might also point to the fact that SOC stocks can vary at field and local scale considerable and thus a very high number of point data is needed to derive statistical properties, that could improve the point-to-grid-cell comparison [see @rammig_generic_2018].f

Using litterfall estimates from LPJmL4, we have been able to estimate the total SOC stocks of the world, which is dominated by SOC under natural vegetation. However, as the world's SOC stock is highly uncertain, which is seen in the wide range of global SOC stock estimates for the first 30 cm of the soil profile [@batjes_harmonized_2016; @hengl_soilgrids250m_2017; @fao_global_2018; @schaphoff_lpjml4_2018-1; @poggio_soilgrids_2021; @sanderman_soil_2017] in Fig. \@ref(fig:SOCglo), the only conclusion we can draw from this is that our result is within a plausible range.

Additionally, we find our $0.2\,\unit{GtC\,yr^{-1}}$ SOC stock change within the period between 1975 and 2010 for the first 30 cm of the soil profile at the upper end of estimates comparing it to estimates of Ito et al. [-@ito_soil_2020] of $0.18\,\pm\,0.41\,\unit{GtC\,yr^{-1}}$ within the period between 1850 and 2014 for whole soil profile. This might be not surprising as the CO~2~ effect is most likely stronger and land-use change effects weaker within our later and shorter modeling period compared to a mean value of the period between 1850 and 2014.
The large uncertainty within estimates of SOC stock and its changes [@ito_soil_2020] again points to the large structural uncertainty within SOC modeling.

To avoid a strong impact of natural land representation and its uncertainties on our results, we focus on SOC changes on cropland. Pristine natural vegetated areas (like permafrost and rain forests) without human land management drop out in our calculation of SOC debt and stock change factors. Natural SOC estimates only influence results when natural land is converted to cropland. Moreover, the temporal dynamic of the SOC debt and stock change factors is not (or only to a small degree) altered by climatic or atmospheric effects on SOC stocks, as they cancel out by taking the difference (for the SOC debt) and ratio (for the stock change factors) of cropland SOC and SOC under hypothetical PNV conditions.

@@

In our model, erosion might however only affect the spatial pattern but not the aggregate SOC pool. As pointed out by Doetterl et al. [-@doetterl_erosion_2016], the final fate of leached or eroded carbon is uncertain and might even offset LUC emissions [@wang_human-induced_2017]. Concurrently, other studies have claimed erosion moves SOC into aquatic reservoirs [@zhang_simulating_2020] and thus changing total global terrestrial SOC stock.
Whereas for soil quality analysis SOC displacement might play an important role, in this budget approach focused especially on the SOC debt, displaced but not emitted SOC can be treated as SOC that remains on the cropland. Erosion and degradation impacts on yields and therefore on soil C inputs are captured by our method as we base them on FAO statistics of actual production. Yet the distribution of production below the country level - which we allocate proportional to LPJmL production potentials that do not reflect erosion feedback to yields - will overestimate yields and therefore biomass inputs to eroded soils.

In comparison with default stock change factors of the IPCC guidelines, our model estimates a stronger decline of SOC stocks (Table \@ref(tab:SCFglo)) for almost all regions. Tropical soils might suffer from low C input rates due to large yield gaps (Global Yield Gap and Water Productivity Atlas. Available URL: www.yieldgap.org (accessed on: 03/01/2022)) and high shares of residue removal and burning in lower-income countries [@smil_crop_1999; @williams_influence_1997; @jain_emission_2014]. Yet, even when comparing our estimates to the low-input stock change factors of the IPCC, our SOC loss is roughly twice as large as the revised 2019 IPCC default values [-@ogle_cropland_in_ipcc_2019], while it shows good agreement with the older default values from IPCC [-@lasco_cropland_in_ipcc_2006]. However, the revised estimates of the IPCC included much more and more recent data points, calling for a closer look on causes for the large deviations between our results and the refined Tier 1 factors. 
On the one hand, our approach does not account for unharvested carbon inputs from weeds or biomass cover on short-term fallows. Shifting agriculture with fallow periods might be prominent in tropical regions. While long-term fallow land (>4 years) is excluded from FAOSTAT as cropland, short-term fallow is not. Thus, our carbon inputs for these areas might be underestimated, leading to too low stock changes factors.
On the other hand, older studies by Don et al. [-@don_impact_2011] estimated SOC losses for tropical soils of around 25% on average corresponding to a stock change factor of 0.75, but also reported a wide range of measured SOC changes from -80% to +58%. Fujisaki et al. [-@fujisaki_forest_2015] however found much lower loss rates of around 9%, attributing the difference to the different time period length since the conversion to cropland. As our results do not specifically account for cropland age and most of the cropland is older than 20 years (as assumed for the default IPCC Tier 1 stock change factors) our stock change factors have to be lower by definition following the steady-state assumption that cropland will continue to approach a new equilibrium. 
For the same reason, our estimates for temperate regions might be lower than both IPCC [-@lasco_cropland_in_ipcc_2006] and IPCC [-@ogle_cropland_in_ipcc_2019] default values.
With the production-increasing impact of irrigation and fertilization on carbon-poor dryland soils, SOC under cropland can also be higher than under PNV with stock change factors above 1 (see Fig. \@ref(fig:SOCmaps)(d)), but these areas are much smaller than where the stock change factors are well below unity.

@@

The effects of agricultural productivity on cropland SOC dynamics, including historical yield trends and associated increases in residue inputs, can be directly accounted for in our modeling approach. In contrast, process-based studies [@pugh_simulated_2015; @herzfeld_soc_2021] often lack data on relevant management aspects that drive production increases. Herzfeld et al. [-@herzfeld_soc_2021] also consider historical management trends for fertilizer and manure inputs as well as on residue removal rates and tillage systems, but cannot reproduce the substantial increase in agricultural productivity over the last decades. Still, they find that compared to no-tillage systems, residue management has much larger potential to affect the strength of their projected future global cropland SOC decline. This is consistent with our finding that increasing SOC inputs from above-ground residues had the strongest effect on the slowing-down of the SOC debt increase (Fig. \@ref(fig:SOCscen)). In line with this, Dangal et al. [-@dangal_improving_2022] finds that no-tillage has only minor impacts on SOC dynamics across parts of the US.
